Does motif participation provide unique information about species risk of extinction?

### 2021-09-28 manuscript/reorder_main_text.tex
Updates from Overleaf

\title{Do motif roles provide unique information about species risk of extinction?}

@@

\section*{Abstract}

    - can we condense omnivory, in-degree any further?
    - - at least omnivory
    - can we shorten any more? Nice to save some money?
    - glossary
    - refs
    - ending

    - repeat network size analyses with \# basal resources - maybe adjust Fig. 1 to reflect n(basal) and C instead of C:S?
    - Fig. 4 not much in discussion, could be a candidate for the SI? Or add more discussion about apparent competition, etc. in discussion sections per motif

@@

    %\subsection*{1 Why are bottom-up effects important for understanding stability?}
     The plant community is the foundation upon which the myriad of species in terrestrial food webs depend for their survival \citep{}. Disturbances in the plant community have been shown to affect important ecosystem properties such as primary \citep{} and secondary production \citep{borer2012plant}, soil respiration and carbon cycling \citep{chen2019plant} and consumer diversity \citep{scherber2010bottom, Baiser2016}. The stability of the plant community is crucial for sustaining a healthy energy flow in the ecosystem as a whole \citep{Rosenblatt2016} and the structure of the plant community promotes stability across all different levels of ecosystem organization \citep{proulx2010diversity,scherber2010bottom}. Loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers, e.g. herbivores, predators, and parasitoids \citep{scherber2010bottom}.
     %This has been shown in both theoretical work~\citep{} and empirical experiments ~\citep{} and results from the vertical propagation of disturbances through the network of interacting species.

     %\subsection*{2 Bayesian networks are a fantastic way to model bottom-up effects}
    The study of how an initial population decline or extinction leads to loss of other species via direct and indirect effects (i.e., secondary extinctions) is a vibrant area of research \citep{curtsdotter2011robustness, dunne2009cascading, Eklof2006}. Traditionally, there are two main approaches for studying secondary extinctions. First, there are topological models, which are solely based on food web structure \citep{dunne2009cascading}. Here, initial extinctions only affect other species in the network once a consumer has lost all of its prey and therefore must go extinct. The second approach uses dynamical models, which explicitly simulate population dynamics using a system of differential equations \citep{binzer2011susceptibility}. Dynamical models take changes in prey or predator densities into account when calculating the densities of other species in the network. These models therefore describe complex effects such as indirect interactions and are generally more realistic than topological models. However, dynamical models are parameter intensive, can only include a limited number of species, and simulations are time-consuming.

    A middle‐ground approach for simulating secondary extinctions are Bayesian networks \citep{Eklof2013}. In this framework, a consumer's probability of extinction depends on the fraction of resources lost and on a baseline probability of extinction which captures causes unrelated to the network structure (e.g., disease, stochastic extinction of small populations). Modelling Bayesian networks is fast and computationally efficient, allowing analysis of networks with high species richness \citep{Haussler2020}. 
    As consumers do not affect the extinction probabilities of their resources, Bayesian networks can only capture bottom-up effects of disturbances within networks. However, Bayesian networks do capture the majority of secondary extinctions captured by dynamical models \citep{Eklof2013}.
    This emphasizes the importance of bottom-up effects in food webs and makes Bayesian networks the ideal framework to use when studying them.

    %\subsection*{3 Global scale network properties are known to affect stability}
    The global structural properties of ecological networks have implications for their functioning \citep{Petchey2002}, stability \citep{Allesina2012} and robustness to disturbances \citep{Dunne2002d, Eklof2006}. For example, more highly-connected networks are more resistant to secondary extinctions \citep{Dunne2002d, Eklof2006} and a modular organization, where some groups of species are tightly interconnected with limited connections between groups, can limit the propagation of disturbances \citep{}.
    %\subsection*{4 meso-scale nw propoerties also affect stability}
    Structural properties on the local scale, i.e. properties of single species, are also well-studied and there is consensus that species with high trophic levels, few prey (low in-degree), and low abundances are generally at higher risks of secondary extinction \citep{binzer2011susceptibility}. However, in-degree and trophic level are both relatively coarse descriptions of a species' role within its food-web~\citep{Cirtwill2018FoodWebs}. 
    For example, specialists may not necessarily have a high risk of extinction if it consumes a basal resource which is at low risk of being affected of disturbance. On the other hand, a consumer with low trophic level may still be at high risk of extinction if it consumes preys which themselves possess high risk of being affected by disturbance.

    %\subsection*{5 Motif participation may similarly affect species' particular extinction risk}
    One way to incorporate more nuance into descriptions of how species fit into their communities is to define species' roles based on their participation in different motifs -- unique combinations of small numbers of interacting species~\citep{Stouffer2007,Stouffer2012}. Importantly, these `motif roles' capture information about species' direct and indirect interactions, incorporating larger-scale network structure into species-level descriptions. 
    This means that two species with the same in-degrees and trophic levels can still participate in different sets of motifs if, for example, one interacts with specialists and the other interacts with generalists~\citep{Cirtwill2018FoodWebs}. 
    This extended description of species structural roles may give additional information of extinction risk or synthesize the information provided by multiple other measures of a species' place in its community.
    At the network level, the frequencies of different motifs are associated with community stability \citep{prill2005dynamic, bascompte2005simple} and with changes to the composition of plant communities~\cite{giling2019plant}. 
    We therefore expect that the motif structure of a network and species' participation in that structure should be related to risks due to disturbance to basal resources.

           %\subsection*{Here we test...}
    Specifically, we are interested in whether a consumer's extinction risk depends on 1) the motif profile of the full food web and 2) the motifs in which a focal species participates. 
    In addition, we want to understand how the information provided by motifs is related to classical global and local food web properties in order to better integrate different aspects of network theory.
    Of the 13 three-species motifs that can occur in empirical food webs there are four motifs that are over-represented in empirical networks and are thought to contribute to stability \citep{Stouffer2007, Borrelli2015a, giling2019plant}: the three-species chain, omnivory, apparent competition and direct competition motifs.
    These four motifs are also the ones which can occur in the acyclic networks required for Bayesian network simulations~\citep{Eklof2013}. 
    Therefore, we focus on these four particularly important motifs.

   % \subsection*{Other, simpler measures of species' roles can also affect extinction risk, and may correlate with motif participation}

    	%Species with high trophic levels may be more likely to go extinct since they depend upon the stability of relatively long food chains~\citep{}. 
        %For the same reasons, the number of motifs a species participates in will also tend to be larger in larger and more-connected networks.
    	%Likewise, the range of possible trophic levels increases with network size, although higher connectance may keep trophic levels low as long as the shortest food chain carries more weight (as in prey-averaged and shortest path trophic levels). 
        %A species' trophic level in turn may affects a species' motif role by restricting the set of positions in which it can appear~\cite{Cirtwill2018EcolLett}. 
    	%To unravel to what extent species motif roles gives additional information on species' probabilities of going extinct, we therefore need to take network structure and other role definitions into account.

@@

		We generated a realistic set of simulated networks based on the niche model~\citep{Williams2000,Stouffer2007} using the function "nichemodel" within the Julia~\citep{Julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw,Delmas2017}. 
		To capture a range of plausible network architectures, we simulated networks with sizes ranging from 50 to 100 species (in steps of 10) and connectances ranging from 0.02 to 0.18 (in steps of 0.04). 
		For full details, see~\citet{Cirtwill2021_inprep}.
        We then filtered these simulated networks to remove any with unreasonable structures.
		Specifically, we removed any network containing disconnected components (species or groups of species not connected to the rest of the network) or any network where the shortest path between a consumer and the closest basal resource was \textgreater6 (as these high trophic levels are very uncommon in empirical food webs ~\citep{}).
		New networks were simulated to replace any removed networks, and this process was repeated until we obtained 100 suitable networks in each combination of size and connectance.

		A Bayesian food web depicts probabilistic relationships among a set of species where each species' probability of persistence depends upon the probabilities of its resource species persisting~\citep{Jensen_Nielsen,Eklof2013}. 
		Therefore, all persistence calculations follow a strict bottom-up routine beginning by determining the status of the primary producers (who do not depend on other species) and continuing on to primary consumers (who depend only on primary producers), and so on up the network.

		While the niche-model food webs can contain cycles (e.g., species A eats B, B eats C, and C eats A), such cycles make it impossible to calculate persistence in the Bayesian network framework~\citep{Tarjan1972}. We therefore removed any cycles within each network.
		To render networks acyclic, we followed~\citet{Allesina2009} and broke cycles by removing links which do not contribute to the robustness of the food web.
		This is achieved by first finding the set of resources for each consumer and then removing the consumer-resource connections which have the lowest eigenvalue centrality~\citep{Allesina2009}.
		These links have the least effect on the overall stability of the network; removing them creates an acyclic network with very similar properties to the original network.
		Next, we order the species from lowest to highest trophic level using a topological sorting routine following \citep{Tarjan1972, Allesinaetal2005}, ensuring that probability calculations follow a strict bottom-up order. 
        After these steps, calculations of persistence probabilities can begin.

@@

		Calculating $P(i)$ or $P(\lnot i)$ is less straightforward when a consumer $i$ loses a fraction $k/n = f$ of its prey. 
		Plausible ecological responses to losing some prey include i) a topological response, where the consumer's probability of extinction is constant unless all resources are lost, ii) a linear response, in which the extinction probability increases linearly with resources loss, and iii) several nonlinear responses where the probability of extinction increases non-linearly with the fraction resource lost. 
		The extinction probability will differ depending on the response function. 
		We use a sigmoid non-linear functional response of consumers to the loss of resources ($\alpha >1, \beta >1$) which has been shown to most accurately capture the secondary extinctions produced by a dynamical model~\citep{Eklof2013}. 
		We use a cumulative density function of a beta distribution for this purpose:
		\begin{equation}
		\label{betafunc}
		P(\lnot i|f) = \pi_{i} + (1 - \pi_{i}) \frac{B(f;\alpha,\beta)}{B(\alpha,\beta)}
		\end{equation}

		We calculate the fraction of resources lost ($f$) for each consumer $i$, to compute $P(\lnot i|f)$. 
		Persistence is determined by comparing $P(\lnot i|f)$ with a randomly drawn number from a uniform distribution between 0-1. 
		If this random number is lower than $P(\lnot i|f)$, the species is considered extinct. 
		The species' marginal probability of persistence is simply $P(i) = 1-P(\lnot i)$.
		In this manner we calculate persistence probabilities for all species in the network, starting from the bottom and working our way up. 
		Probabilities of extinction for resource species are used when calculating probabilities for higher level species. 
		Some species will be extinct due to primary extinctions (stochastic extinction with all resources left), while others will experience secondary extinctions; as a response to primary extinctions (partial or total loss of resources). 
		As this method is probabilistic, repeating the simulation will yield different results. 
		To minimize the effect of this variation we repeat these calculations 100 000 times per web, with unique random draws when calculating $P(i)$.
		The final persistence for each species is the mean over these runs.

	\subsection*{Defining motif profiles and participation [species roles]}

        As a first measure of the meso-scale structure, we calculated the `motif profile' of each network.
        This is, in our case, a four-dimensional vector of the number of each three-species motif in the network.
        To separate differences in motif structure from differences in network size, we normalized motif profiles by dividing the count of each motif by the total count of all motifs.
        The motif profile for each network therefore describes the proportions of each motif and has a sum of exactly one.

        We then describe meso-scale structure at the species level by defining each species' motif participation and motif roles.
        Motif participation describes the frequency with which a focal species appears in each of the four motifs, analogous to the network motif profile~\citep{Stouffer2012}.
        A species' motif role describes the frequency with which a focal species appears in each \emph{unique position} (e.g., top, middle, or bottom species in the three-species chain, top or bottom species in the apparent competition motif) within each motif ~\citep{Stouffer2012,Cirtwill2017}.
		We calculated species' motif participation and roles using the Python package \emph{pymfinder}~\citep{pymfinder}.
        As with motif profiles of networks, we normalized participation (and role) vectors by dividing each count by the total number of motifs (motif positions) the species appears in.
        Note, however, that this normalization does \emph{not} control for differences due to degree \emph{per se} as high-degree species also tend to participate in different types of motifs than low-degree species~\citep{Cirtwill2021_inprep}.

        Finally, we calculated simpler measures of a species' role within its community: degree and trophic level.
		As the Bayesian network approach only accounts  bottom-up effects, out-degree (number of predators) is irrelevant to extinction risk.
		We therefore focused on species' in-degrees (number of prey).
		Note, however, that out-degree may also be important in topological simulations which include both top-down and bottom-up effects.

		There are many definitions of trophic level available, most of which tend to be very similar~\citep{Carscallen2012}.
		Here we use shortest trophic level (STL): the length of the shortest food chain between the focal species and any basal resource~\citep{Williams2004}. 
		Basal resources have an STL of one, any species consuming a basal resource has an STL of two, etc.
        STL is therefore a measure of the minimum number of steps away a species is from any disturbance to the base of a food web.

	\subsection*{Statistical analysis}

        \subsubsection*{How does species persistence vary with motif participation?}

            We are interested in whether participating more often in a certain motif (e.g., direct competition) might increase persistence.
            % (see \emph{Appendix SXX} for relationships between persistence and counts of motifs).
            To test this, we fit four linear mixed-effect models (LMMs; one per motif included in the Bayesian networks).
            These models included the proportion of the species' role made up by the focal motif, the level of disturbance to basal species, and the interaction between the two.
            We centered and scaled the predictors before fitting the LMMs.
            The models also included a random intercept for the interaction between species richness and connectance (equations.
            All models were fit using the R~\citep{R} function `lmer' from the package \emph{lmerTest}~\citep{lmerTest}.
            We calculated marginal (fixed effects only) and conditional (fixed and random effects) $R^2$ using the R~\citep{R} function `r.squaredGLMM' from the package \emph{MuMIn}~\citep{MuMIn}.

@@

        \subsubsection*{How does motif participation vary with degree and trophic level?}

            A further complication when trying to understand whether relationships between motifs and probability of extinction depend upon underlying relationships between degree or trophic level and extinction 
            We also fit linear mixed-effect models relating the count or proportion of each motif to degree or trophic level and a random effect of the interaction between network size and connectance using the R~\citep{R} function `lmer' from the package \emph{lmerTest}~\citep{lmerTest}.
            We performed similar analyses to test whether a species' motif role is related it in-degree or trophic level (\emph{Appendix SI}).

    	\subsubsection*{How consistent are motif-persistence relationships?}

    		As well as identifying general relationships between motif participation and persistence, we were also interested in whether these general trends were consistent across networks with different global properties and different levels of disturbance. 
            To test this, we performed a linear regression of persistence against motif participation within each simulated network, for each level of disturbance to basal resources.
            These regressions were fit using the R~\citep{R} base function `lm'.
            We then analyse the distribution of slopes for the effect of motif participation across levels of disturbance, network size, and connectance. 
            We performed similar analyses to test whether appearing in different positions in each motif (i.e., the motif role) is related to a species' persistence (\emph{Appendix SH}).

        \subsubsection*{How does network mean persistence vary with motif profiles?}

            To test whether networks with similar motif profiles tend to have similar mean persistences among consumers, we fit a PERMANOVA relating Bray-Curtis dissimilarity in motif profiles to the mean persistence of all consumers in a network (averaged across all levels of disturbance to basal resources).
            We then tested whether dispersion was homogeneous across levels of persistence using an ANOVA and whether there was a linear relationship between persistence (rounded to three decimal places) and variability of motif profiles using a linear regression.
            In case high or low mean persistences are associated with more variable network structure, we calculated dispersion of motif structure for each level of mean persistence (rounded to three decimal places to allow multiple networks to have the same value of mean consumer persistence and treated as a categorical variable). 
            All statistical tests were conducted as when relating motif profiles to global network structure.

            To supplement these overall tests, we also fit four regressions testing whether persistence had a linear relationship to the proportion of a network profile made up of each motif, and whether this varied with the level of disturbance.
            Each regression included the proportion of a motif, disturbance, and their interaction as predictors.
            All regressions were fit using the R~\citep{R} base function `lm'.

        \subsubsection*{How do network motif profiles vary with global structure?}

            As well as affecting persistence directly, global network structure (i.e., size and connectance) may affect the meso-scale structure of the network.
            This is especially likely since we include only networks which can retain all initial species after 1000 rounds of simulated Lotka-Volterra population dynamics~\citep{Cirtwill2021_inprep}:  some meso-scale structures might be more likely to allow all species to persist than others in networks with different size and connectance constraints.
            We test whether networks with the the same size and connectance have more similar motif profiles using a PERMANOVA~\citep{Anderson2001} of Bray-Curtis dissimilarity among motif profiles against network size, connectance, and their interaction.
            Significance was calculated based on 9999 permutations.
            PERMANOVA tests may give false positive results if variability is not homogeneous among groups (i.e., levels of network size or connectance).
            To test whether this applies in our case, we fit an ANOVA of the dispersion of motif profiles about the centroid for each combination of network size and connectance. 
            We fit the PERMANOVA using the function `adonis', calculated dispersion using the function `betadisper', and fit the ANOVA using the function `anova', all from the R~\citep{R} package \emph{vegan}~\citep{vegan}.
            To gain a more detailed picture of how the proportion of each motif may vary with global network structure, we also fit four regressions (one per motif) relating the proportion of a motif to network size, connectance, and their interaction.
            We fit each regression using the R~\citep{R} base function `lm'.

        \subsubsection*{How does network mean persistence vary depending on network properties?}

            Both species richness (S) and connectance (C) are known to influence persistence in both simulation studies and empirical food webs.
            In addition, increasing the probability of extinction for basal species ($\pi_{disturbed}$ above) is intended to decrease persistence of species at higher trophic levels.
            Although these effects are not the main focus of our manuscript, we must establish these background effects before measuring the effects of meso-scale structure on persistence.
            The persistence of basal species in our simulations is determined only by their baseline probability, being either $\pi_{base} = 0.1$ or higher, dependent on $\pi_{disturbed}$.
            We therefore considered only non-basal species (i.e., those with at least one prey) throughout this manuscript.

            To test for effects of global network properties, we fit a linear regression relating consumer species' probabilities of persistence to network size, connectance, the level of disturbance to basal resources, and all interactions between them. 
            All predictors were centered and scaled before fitting the model. 
            As the three-way interaction was significant, we did not simplify the model.
            All models were fit using the R~\citep{R} base function `lm'.

        \subsubsection*{How does persistence vary with degree and trophic level?}

            A species' motif participation is likely related to its degrees and trophic level since all three are measures of network position.
            To understand how this may have affected the results above, we first fit two linear mixed-effects regressions relating persistence to degree or trophic level, $\pi_{disturbed}$, and their interaction.
            As above, we centered and scaled the predictors.
            We also included a random effect of the interaction between network size and connectance in order to account for the differences in persistence between different networks. 
            These regressions provide context about how persistence varies with simpler measures of species' roles within networks.

\section*{Results}

    We are interested in how a species' participation in different motifs affects its persistence, and how these relationships relate to other measures of network structure. 
    We analyse this for different levels of disturbance to the basal level. 
    To provide context, we also investigate relationships between persistence and classic global (size and connectance) and local (in-degree and trophic level) network properties.
    % The results are organized as follows: 1) How network persistence varies with network motif distribution. 2) How network persistence varies with global structural properties. 3) How network motif distribution varies with global structural properties. 4) How species probability of persistence varies with its motif profile. 5) How species persistence varies with its characteristics such as TL and in-degree. 6) How a species motif profile depend on its characteristics. 7) How a species persistence varies with its motif role (position in respective motif).

@@

       The effects of motif participation on a species' probability of persistence depends on the level of disturbance. When disturbance is low, species' persistence increases with increasing proportions of the omnivory and direct competition motifs. Conversely, a species' persistence decreases with an increasing proportion of the apparent competition motif. Persistence is not significantly related to the proportion of the three-species chain motif (Figure~\ref{fig:prop_lmer_all}).

        At high levels of disturbance, a species' persistence decreases as the proportion of the three-species chain increases, but persistence increases as the proportion of any other motif increases (Figure~\ref{fig:prop_lmer_all}). This means that increased disturbance strengthens the effects of the three-species chain motif and changes the direction of the effect of the apparent competition motif  (Table S4, \emph{Appendix SD}). The effect of omnivory on persistence is still positive but weaker at higher levels of disturbance, and the relationship between persistence and direct competition was largely unaffected by disturbance.

@@

>>>>

\title{Does motif participation provide unique information about species risk of extinction?}

@@

\section*{Abstract}
    Bottom-up effects of disturbance to basal resources can strongly affect the persistence of consumers.
    However, predicting which species will be most affected by increasing disturbance is not straightforward.
    In-degree and trophic level are known to affect a consumer's vulnerability to bottom-up disturbance, but these are relatively coarse measures of a species' place in a food web.
    Here we explore how roles based on participation in three-species motifs are related to persistence, and how these motif-based roles relate to simpler measures.
    We show that a consumer species' frequency of participation in different motifs is related to its probability of persistence. Similarly, the overall motif profile of a network was related to the average persistence of consumer species.
    Notably, we found that networks containing a higher proportion of the `omnivory' motif tend to have lower mean persistence following disturbance to basal species, while more frequent participation in the omnivory motif was associated with higher probability of persistence.
    Additionally, we find that relationships between motif participation and persistence can synthesize the information provided by in-degree and trophic level.
    Motif participation therefore provides a relatively rich description of a species' structural role in its community and captures important trends in vulnerability to extinction.

    Anna AA: The whole article is now approximately 4400 words (not counting fig texts); intro 850, methods 1600, results 620, discussion 1300. See discussion on section statistical analysis in the methods, I think we can make it shorter
    - glossary
    - refs
    - ending

    - Fig. 4 not much in discussion, could be a candidate for the SI? Or add more discussion about apparent competition, etc. in discussion sections per motif

@@

    %\subsection*{1 Why are bottom-up effects important for understanding stability?}
     The plant community is the foundation upon which the myriad species in terrestrial food webs depend for their survival. Disturbances in the plant community have been shown to affect important ecosystem properties such as primary \citep{Hector1999} and secondary production \citep{borer2012plant}, soil respiration and carbon cycling \citep{chen2019plant} and consumer diversity \citep{scherber2010bottom, Baiser2016}. The stability of the plant community is crucial for sustaining a healthy energy flow in the ecosystem as a whole \citep{Rosenblatt2016} and the structure of the plant community affects stability across all levels of ecosystem organization \citep{proulx2010diversity,scherber2010bottom}. Loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers: herbivores, predators,  parasitoids, etc. \citep{scherber2010bottom}.

     %\subsection*{2 Bayesian networks are a fantastic way to model bottom-up effects}
    How population decline or extinction of one species leads to loss of other species via direct and indirect effects (i.e., secondary extinctions) is a vibrant area of research \citep{curtsdotter2011robustness, dunne2009cascading, Eklof2006}. Traditionally, there are two main approaches for studying secondary extinctions. First, there are topological models, based only on food web structure \citep{dunne2009cascading}. Here,  extinctions only affect other species in the network once a consumer has lost all of its prey and therefore must go extinct. The second approach uses dynamical models which explicitly simulate population dynamics using a system of differential equations \citep{binzer2011susceptibility}. Dynamical models take changes in prey or predator densities into account when calculating the densities of other species in the network. 
    This additional detail means that realistic processes such as indirect interactions can be included, but also means that dynamical models are more parameter intensive, and simulations are more time-consuming, than topological models.

    A middle‐ground approach for simulating secondary extinctions is to use  Bayesian networks \citep{Eklof2013}. 
    In this framework, a consumer's probability of extinction depends on the fraction of resources lost and on a baseline probability of extinction which captures causes unrelated to the network structure (e.g., disease, stochastic extinction of small populations).
    Modelling Bayesian networks is fast and computationally efficient, allowing analysis of networks with high species richness \citep{Haussler2020}. 
    As consumers do not affect the extinction probabilities of their resources, Bayesian networks can only capture bottom-up effects of disturbances within networks. 
    However, Bayesian networks do capture the majority of secondary extinctions captured by dynamical models \citep{Eklof2013}.
    This emphasizes the importance of bottom-up effects in food webs and makes Bayesian networks the ideal framework to use when studying them.

    %\subsection*{3 Global scale network properties are known to affect stability}
    Using Bayesian networks, researchers can explore how the structures of realistically-large networks shape their responses to disturbance.
    The global (whole-network) structural properties of ecological networks have implications for their functioning \citep{Petchey2002}, stability \citep{Allesina2012} and robustness to disturbances \citep{Dunne2002, Eklof2006}. 
    For example, more highly-connected networks are more resistant to secondary extinctions \citep{Dunne2002, Eklof2006} and a modular organization, where some groups of species are tightly interconnected with limited connections between groups, can limit the propagation of disturbances \citep{Krause2003, Teng2004}.

    %\subsection*{4 meso-scale nw propoerties also affect stability}
    Structural properties on the local (single-species) scale are also known to affect response to disturbance. 
    Species with high trophic levels (long paths to basal resources), low in-degree (few prey), and low abundances are generally at higher risks of secondary extinction \citep{binzer2011susceptibility}. 
    However, in-degree and trophic level are both relatively coarse descriptions of a species' role within its food-web~\citep{Cirtwill2018FoodWebs}. 
    For example, a specialist may not necessarily have a high risk of extinction if it consumes a basal resource which is unlikely to be affected by disturbance. 
    On the other hand, a consumer with low trophic level may be at high risk of extinction if it consumes prey which are themselves very likely to be affected by disturbance.

    %\subsection*{5 Motif participation may similarly affect species' particular extinction risk}
    One way to incorporate more nuance into descriptions of how species fit into their communities is to define species' roles based on their participation in different motifs -- unique combinations of small numbers of interacting species~\citep{Stouffer2007,Stouffer2012}. Importantly, these `motif participation roles' capture information about species' direct and indirect interactions, incorporating larger-scale network structure into species-level descriptions~\citep{Cirtwill2015JAE}. 
    This means that two species with the same in-degrees and trophic levels can still participate in different sets of motifs if, for example, one interacts with specialists and the other interacts with generalists~\citep{Cirtwill2018FoodWebs}. 
    This extra detail  may be useful when predicting  extinction risk since, at the network level, the frequencies of different motifs are associated with community stability \citep{prill2005dynamic, bascompte2005simple} and with changes to the composition of plant communities~\cite{giling2019plant}. 
    We therefore expect that the motif structure of a network and species' participation in that structure should be related to risks due to disturbance to basal resources.

           %\subsection*{Here we test...}
    Specifically, we are interested in whether a consumer's extinction risk depends on 1) the motif profile of the full food web and 2) the motifs in which a focal species participates. 
    In addition, we want to understand how the information provided by motifs is related to classical global and local food web properties in order to better synthesize information provided by different scales of network structure.
    Of the 13 three-species motifs that can occur in empirical food webs, there are four motifs that are over-represented in empirical networks and are thought to contribute to stability \citep{Stouffer2007, Borrelli2015a, giling2019plant}. These four motifs, the three-species chain, omnivory, apparent competition, and direct competition, are also the only ones which can occur in the acyclic networks required for Bayesian network simulations~\citep{Eklof2013}. 
    Therefore, we focus on these four particularly important motifs.

@@

		We generated a realistic set of simulated networks based on the niche model~\citep{Williams2000,Stouffer2007} using the function "nichemodel" within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw}. 
		To capture a range of plausible network architectures, we simulated networks with sizes ranging from 50 to 100 species (in steps of 10) and connectances ranging from 0.02 to 0.18 (in steps of 0.04). 
		For full details, see~\citet{Cirtwill2021_inprep}.
        We then filtered these simulated networks and removed any network containing disconnected components
        % (species or groups of species not connected to the rest of the network) 
        or any network where any species had a shortest trophic level \textgreater6 as such high trophic levels are very uncommon in empirical food webs ~\citep{Riede2011}.
		New networks were simulated to replace any removed networks, and this process was repeated until we obtained 100 suitable networks in each combination of size and connectance.
		All networks were rendered acyclic following~\citet{Allesina2009} and topologically sorted from lowest to highest trophic level (Appendix\emph{S1}).

		A Bayesian food web depicts probabilistic relationships among a set of species where each species' probability of persistence depends upon the probabilities of its resource species persisting~\citep{Jensen_Nielsen,Eklof2013}. 
		All persistence calculations therefore begin  by determining the status of the primary producers (who do not depend on other species).
		Calculations continue in a strictly bottom-up manner with primary consumers (who depend only on primary producers), and so on up the network.

		While niche-model food webs can contain cycles (e.g., species A eats B, B eats C, and C eats A), such cycles make it impossible to calculate persistence in the Bayesian network framework~\citep{Tarjan1972}. 
		We therefore removed any cycles within each network by removing the link in each cycle which contributed least to the robustness of the food web, following~\citet{Allesina2009}.
		%This is achieved by first finding the set of resources for each consumer and then removing the consumer-resource connections which have the lowest eigenvalue centrality~\citep{Allesina2009}.
		%These links have the least effect on the overall stability of the network; removing them creates an acyclic network with very similar properties to the original network.
		Next, we ordered the species from lowest to highest trophic level using a topological sorting routine following \citep{Tarjan1972, Allesinaetal2005}, ensuring that probability calculations follow a strict bottom-up order. 
        After these steps, calculations of persistence probabilities can begin.

@@

		Calculating $P(i)$ or $P(\lnot i)$ is less straightforward when a consumer $i$ loses a fraction $k/n = f$ of its prey. 
		We use a sigmoid non-linear functional response of consumers to the loss of resources, which has been shown to most accurately capture the secondary extinctions produced by a dynamical model~\citep{Eklof2013}. For details see SI.
		We calculate the fraction of resources lost ($f$) for each consumer $i$, to compute $P(\lnot i|f)$. 
		Persistence is determined by comparing $P(\lnot i|f)$ with a randomly drawn number from a uniform distribution between 0-1. 
		If this random number is lower than $P(\lnot i|f)$, the species is considered extinct. 
		The species' marginal probability of persistence is simply $P(i) = 1-P(\lnot i)$.
		In this manner we calculate persistence probabilities for all species in the network, starting from the bottom and working our way up. 
		Probabilities of extinction for resource species are used when calculating probabilities for higher level species. 
		Some species will be extinct due to primary extinctions (stochastic extinction with all resources left), while others will experience secondary extinctions; as a response to primary extinctions (partial or total loss of resources). 
		Although it is possible to solve the Bayesian network exactly \citep{Eklof2013}, this is cumbersome for larger networks. As numerical simulations are highly efficient and produce the same result \citep{Haussler2020} we use that method here. 
		As this method is probabilistic, repeating the simulation will yield different results. 
		To minimize the effect of this variation we repeat these calculations 100 000 times per web, with unique random draws when calculating $P(i)$.
		The final persistence for each species is the mean over these runs.

	\subsection*{Defining motif profiles and species roles}

        We calculated the `motif profile' of each network. This is, in our case, a four-dimensional vector of the number of each three-species motif in the network. To separate differences in motif structure from differences in network size, we normalized motif profiles by dividing the count of each motif by the total count of all motifs. The motif profile for each network therefore describes the proportions of each motif and has a sum of exactly one.

        We then defined each individual species' motif participation. Motif participation describes the frequency with which a focal species appears in each of the four motifs, analogous to the network motif profile~\citep{Stouffer2012}.

		We calculated species' motif participation using the Python package \emph{pymfinder}~\citep{pymfinder}.
        As with motif profiles of networks, we normalized participation (and role) vectors by dividing each count by the total number of motifs (motif positions) the species appears in.
        Note, however, that this normalization does \emph{not} control for differences due to degree \emph{per se} as high-degree species also tend to participate in different types of motifs than low-degree species~\citep{Cirtwill2021_inprep}.

        Finally, we calculated simpler measures of a species' role within its community: in-degree and trophic level (shortest trophic level (STL): the length of the shortest food chain between the focal species and any basal resource~\citep{Williams2004}).

	\subsection*{Statistical analysis} 
	We performed multiple statistical analysis, investigating 1) how species persistence vary with motif participation (Appendix XX), 2) how motif participation vary with degree and trophic level (Appendix XX), 3) how network mean persistence vary with network motif profiles (Appendix XX), 4) how network motif profiles vary with global structural properties (Appendix XX), 5) how network mean persistence vary depending on network properties (Appendix XX), and finally 6) how persistence varies with degree and trophic level (Appendix XX). The analyses are briefly described below. For detailed descriptions see Appendix XX.

    \subsection*{AE - Statistical analysis}

            Effects on persistence from global network properties (i.e. $S$ and $C$) must be established before measuring the effects of meso-scale structure on persistence. Therefore, we fit a linear regression relating consumer species' probabilities of persistence to network size, connectance, the level of disturbance to basal resources, and all interactions between them. As the three-way interaction was significant, we did not simplify the model.

            Global network properties may affect the meso-scale structure of the network. Therfore we test whether networks with the the same size and connectance have more similar motif profiles using a PERMANOVA~\citep{Anderson2001} of Bray-Curtis dissimilarity among motif profiles against network size, connectance, and their interaction. For details see SI(XX).

            To test whether networks with similar motif profiles tend to have similar mean persistences among consumers, we fit a PERMANOVA relating Bray-Curtis dissimilarity in motif profiles to the mean persistence of all consumers in a network (averaged across all levels of disturbance to basal resources).  For details see SI(XX).

            To supplement these overall tests, we also fit four regressions testing whether persistence had a linear relationship to the proportion of a network profile made up of each motif, and whether this varied with the level of disturbance.
            Each regression included the proportion of a motif, disturbance, and their interaction as predictors.

            To test if participating more often in a certain motif affect a species persistence, we fit four linear mixed-effect models (LMMs; one per motif included in the Bayesian networks).
            These models included the proportion of the species' role made up by the focal motif, the level of disturbance to basal species, and the interaction between the two. The models also included a random intercept for the interaction between species richness and connectance.

@@

            To test whether these general trends were consistent across networks with different global properties and different levels of disturbance, we performed a linear regression of persistence against motif participation within each simulated network, for each level of disturbance to basal resources. We then analyse the distribution of slopes for the effect of motif participation across levels of disturbance, network size, and connectance.

            A species' motif participation can be related to its in-degree and trophic level. Therefore we fit two linear mixed-effects regressions relating persistence to degree or trophic level, $\pi_{disturbed}$, and their interaction. We also included a random effect of the interaction between network size and connectance in order to account for the differences in persistence between different networks.

            Species' motif participation may covary with measures of network structure and species' position within the network.To establish this context for motif participation, we fit a linear mixed-effect model relating the frequency of participation in each motif to network size, connectance, and their interaction, as for network motif profiles (\emph{Appendix SG}).We performed similar analyses to test whether a species' motif role is related it in-degree or trophic level (\emph{Appendix SI}).

            All predictors were centered and scaled before fitting the models. 
\section*{Results}

    %We are interested in how a species' participation in different motifs affects its persistence, and how the overall motif profile of a network relate to the average persistence of consumer species. We analyse this for different levels of disturbance to the basal level. To provide context, we also investigate relationships between persistence and classic global (size and connectance) and local (in-degree and trophic level) network properties.

@@

       The relationship between frequency of participation in a motif and a species' probability of persistence depended on the level of disturbance. 
       When disturbance was low, species' persistence increased with increasing proportions of the omnivory and direct competition motifs. 
       Conversely, persistence decreased with an increasing proportion of the apparent competition motif. 
       Persistence was not significantly related to the proportion of the three-species chain motif (Figure~\ref{fig:prop_lmer_all}).

        At high levels of disturbance, a species' persistence decreased as the proportion of the three-species chain increased, but persistence increased as the proportion of any other motif increased (Figure~\ref{fig:prop_lmer_all}). 
        This means that increased disturbance strengthened the effects of the three-species chain motif and changed the direction of the effect of the apparent competition motif  (Table S4, \emph{Appendix SD}).
        The relationship between omnivory and persistence was still positive but weaker at higher levels of disturbance, and the relationship between persistence and direct competition was largely unaffected by disturbance.

@@

    \subsection*{Motif participation and other network properties}

       Species' motif participation also varied with their in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}).
       Species with higher in-degree tended to have higher frequencies of participation in the omnivory motif. Proportions of the other three motifs tended to decline with increasing degree.
       Species with higher trophic levels tended to have higher proportions of apparent competition and three-species chains.
